\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-10-24 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
add NSERC DG in acknowledgment

options(kableExtra.latex.load_packages = FALSE)
library(casebase)
library(survival)
library(splines)
library(ggplot2)
library(colorspace)
library(cowplot)
library(data.table)
library(glmnet)
library(kableExtra)
library(magrittr)
library(pracma)
library(flexsurv)
library(prodlim)
# library(riskRegression)
library(tibble)
library(visreg)
library(lubridate)
library(tidyverse)
library(scales)
options(digits = 2)
# set the seed for reproducible output
set.seed(1234)

# Turn on/off sections for faster compilation
eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- TRUE
eval_cs4 <- FALSE
eval_colophon <- FALSE # Only for draft
echo_plot_code <- TRUE
# Define theme for plots
paper_gg_theme <- ggplot2::theme_bw(base_size = 10) + ggplot2::theme(legend.position = "bottom")
# Define colour palette for CS 3
q4 <- colorspace::qualitative_hcl(4, palette = "Dark 3")
names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
```

@@

Several authors have since pursued fully parametric approaches that made the 
fitting of smooth survival curves much more transparent and intuitive through 
generalized linear models. The key feature of these procedures is splitting the 
time axis into discrete intervals. Whitehead [-@whitehead1980fitting] showed the 
equivalence between a Cox model and a Poisson regression with a parameter for 
each event time; Carstensen [-@copenhagen2012needs] provides a nice exposition of this 
equivalence with a real data example and supporting R code for computing 
standard errors. Arjas & Haara [-@arjas1987logistic] and Efron [-@efron1988logistic] treated each 
patient-day as a Bernoulli random variable with probability equal to the 
discrete hazard rate. A potential issue with these approaches is that the 
number of time bins need to be chosen by the data analyst. On the one hand, a fine grouping 
of times may result in few (or none) events per interval, which then leads 
to instability in the Newton-Raphson procedure for estimation 
[@kalbfleisch2011statistical, Section 4.8]. 
On the other hand, a coarse grouping could potentially mask 
nonlinear trends in the hazard function.

Rather than discretizing time, Hanley & Miettinen [-@hanley2009fitting] 
selected a discrete set of person-time coordinates (`person-moments') in 
continuous time from all observed follow-up experience constituting the 
study base. By doing so, they obtained a likelihood expression for the hazard 
function that is equivalent to that of logistic regression. More specifically, all person-moments when the event of interest 
occurred are selected as the case series, complemented by a randomly sampled 
base series of person-moments serving as controls. This approach allows 
flexible modeling of the hazard function by including time as a covariate 
(e.g. using splines or general additive models). Furthermore, time-dependent 
covariates can be modeled through interactions with time. In short, Hanley & 
Miettinen [-@hanley2009fitting] use the well-understood logistic regression 
for directly modeling the hazard function, without requiring a discrete-time model.

@@

```{r erspc-data-mutate, eval = eval_cs1, echo = FALSE}
data("ERSPC")
ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
                       levels = c(0,1), 
                       labels = c("Control group", "Screening group"))
```

```{r erspc-data, eval = eval_cs1, echo=TRUE}
pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
                               event = "DeadOfPrCa", exposure = "ScrArm")
inherits(pt_object, "popTime")
attr(pt_object, "exposure")
```

@@

For our first case study, we return to the ERSPC study and investigate the differences in risk between the control and screening arms. 
Previous re-analyses of these data suggest that the 20% reduction in prostate cancer death due to screening was an underestimate [@hanley2010mortality]. 
The estimated 20% (from a proportional hazards model) did not account for the delay between screening and the time the effect is expected to be observed. 
As a result, the null effects in years 1–7 masked the substantial reductions that began to appear from year 8 onward. 
This motivates the use of a time-dependent hazard ratio which can easily be fit with the \pkg{casebase} package by including an interaction term with time in the model. 
We fit a flexible hazard by using a smooth function of time modeled with a penalized cubic spline basis with 2 degrees of freedom (implemented in the \code{survival::pspline} function). 
The model is fit using \code{fitSmoothHazard} with the familiar formula interface:

```{r, eval = eval_cs1, echo = -1}
set.seed(1953)
fit <- fitSmoothHazard(DeadOfPrCa ~ pspline(Follow.Up.Time, df = 2) * ScrArm, 
                       data = ERSPC, ratio = 100)
```

Note that we did not have to specify any cut points, as would be the case with the \code{survSplit} function in the \pkg{survival} package. 
The output object from \code{fitSmoothHazard} inherits from the \code{singleEventCB} and \code{glm} classes.
For this reason, we can leverage the \code{summary} method for \code{glm} objects to output a familiar summary of the results:

@@

As noted in the Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models. 
Therefore, we can easily test the significance of the spline term and its interaction with time:

```{r erscp-compare, eval = eval_cs1, echo = TRUE}
anova(fit, test = "LRT")
```

Similarly, to compare different models (e.g. time modeled linearly), we could compute Akaike's Information Criterion (AIC) for each model and compare them.

### Time-dependent hazard ratios

In what follows, the hazard ratio for a variable $X$ is defined as

$$
\frac{\lambda\left(t | X=x_1, \mathbf{Z}=\mathbf{z_1} ; \hat{\beta}\right)}{\lambda(t | X=x_0, \mathbf{Z}=\mathbf{z_0} ; \hat{\beta})}
$$
where $\lambda(t|\cdot;\hat{\beta})$ is the hazard rate as a function of the variable $t$ (which is usually time, but can be any other continuous variable), $x_1$ is the value of $X$ for the exposed group, $x_0$ is the value of $X$ for the unexposed group, $\mathbf{Z}$ are other covariates in the model which are equal to $\mathbf{z_1}$ in the exposed and $\mathbf{z_0}$ in the unexposed group, and $\hat{\beta}$ are the estimated regression coefficients. 
As indicated by the formula above, it is most instructive to plot the hazard ratio as a function of a variable $t$ only if there is an interaction between $t$ and $X$. 
Otherwise, the resulting plot will simply be a horizontal line across time.

The \code{plot} method with \code{type="hr"} for objects of class \code{singleEventCB} can be used to compute time-dependent hazard ratios and confidence intervals. 
In Figure \ref{fig:interaction-ERSPC}, we show the estimated hazard ratio and 95% confidence interval for screening vs. control group as a function of time. 
Note that we must specify the covariate profiles and times for the predicted hazards.

```{r, echo = TRUE, eval = FALSE}
new_time <- seq(1, 12, by  = 0.1)
new_data <- data.frame(ScrArm = factor("Control group",
                                         levels = c("Control group","Screening group")),
                      Follow.Up.Time = new_time)
plot(fit, type = "hr", newdata = new_data,
     var = "ScrArm", xvar = "Follow.Up.Time", ci = T)
```

```{r interaction-ERSPC, eval = eval_cs1, fig.width=8, fig.height=6, fig.cap='Estimated hazard ratio and 95\\% confidence interval for screening vs. control group as a function of time in the ERSPC dataset. Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic pspline basis of follow-up time and treatment arm. 95\\% confidence intervals are calculated using the delta method. The plot shows that the effect of screening only begins to become statistically apparent by year 7. The 25-60\\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\\% reported in the original article.'}
new_time <- seq(1, 12, by  = 0.1)
new_data <- data.frame(ScrArm = factor("Control group",
                                         levels = c("Control group","Screening group")),
                      Follow.Up.Time = new_time)

par(oma = c(0,1,0,8))
tt <- plot(fit,
     type = "hr",
     newdata = new_data,
     var = "ScrArm",
     increment = 1,
     xvar = "Follow.Up.Time",
     ci = T,
     xlab = "Follow-up time (years)",
     ylab = "Death from prostate cancer hazard ratio",
     ylim = c(0,1.50),
     yaxt = 'n',
     xaxt = "n",
     rug = TRUE)

### Hazard functions

Modeling the hazard function directly allows us to easily visualize it with the \code{plot} method and \code{type="hazard"} for objects of class \code{singleEventCB}. 
We plot the hazard functions for both treatment arms in Figure \ref{fig:cs1hazard}. 
The pattern we see is consistent with the population-time plot shown in Figure \ref{fig:plot-erspc-data} C, where more events are observed at later follow-up times. 
The drop at the end can be explained by the fact that very few observations were followed for the entire 15 year period.

```{r cs1hazard, eval=eval_cs1, echo=TRUE}
plot(fit, type = "hazard",
     hazard.params = list(xvar = "Follow.Up.Time",
                          by = "ScrArm"))
```

### Absolute risk

@@

```{r, eval=FALSE, echo=FALSE}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- c(0, 1)

@@

```{r, eval=eval_cs1, echo=TRUE}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0,14,0.1)
risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
head(risk)
```

<!--In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated risk function from \pkg{casebase} on the Cox model estimate. -->

@@

```{r erspc-cox-cif, eval = FALSE, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk functions for control and screening groups in the ERSPC data. We plot the estimate from the Cox model using \\code{survival::survfit} (solid line) and the estimate from case-base sampling (dashed line). We used a cubic B-spline expansion of time."}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
myCols <- cbPalette[c(4,7)]

@@

We can subsequently compute confidence intervals for the risk function using the method \code{confint.absRiskCB}:

```{r eval = eval_cs1, echo = TRUE}
conf_ints <- confint(risk, fit)
head(conf_ints)
```

These point-wise confidence intervals are constructed using parametric bootstrap; the validity of this approach relies on the asymptotic normality of the MLE. 
In Figure \ref{fig:erspc-cif-conf}, we see the 95% confidence bands around the estimates. 
We also overlay the Kaplan-Meier curves as a reference.

```{r erspc-cif-conf, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\\% confidence bands."}

tt <- conf_ints

# plot absrisk and 95% CI
i.backw <- order(new_time, decreasing = TRUE)
i.forw <- order(new_time)

@@

```{r}
knitr::knit_exit()
```

@@

We want to compare our hazard ratio estimates to that obtained from a Cox regression (using the \pkg{survival} package version 3.2-13).

@@

```{r supportData, eval = eval_cs3}
library(riskRegression)
scaleFUNy <- function(x) sprintf("%.2f", x)
scaleFUNx <- function(x) sprintf("%.0f", x)
data(support)
# Change time to years
support$d.time <- support$d.time/365.25

@@

We would like to thank the anonymous reviewers for their insightful comments and criticisms. We would also like to thank Yi Yang for helpful discussions on penalized regression models. \mbox{Bhatnagar} gratefully acknowledges funding via a Discovery Grant from the Natural Sciences and Engineering Research Council of Canada (NSERC).

>>>>

options(kableExtra.latex.load_packages = FALSE)
library(casebase)
library(colorspace)
library(cowplot)
library(data.table)
library(glmnet)
library(kableExtra)
library(magrittr)
library(pracma)
library(flexsurv) 
library(prodlim)
library(riskRegression)
library(splines)
library(survival)
library(tibble)
library(visreg)
library(lubridate)
library(tidyverse)
library(scales)
options(digits = 2)
# set the seed for reproducible output
set.seed(1234)

# Turn on/off sections for faster compilation
eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- TRUE
eval_cs4 <- FALSE
eval_colophon <- FALSE # Only for draft
echo_plot_code <- TRUE
# Define theme for plots
paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
# Define colour palette for CS 3
q4 <- qualitative_hcl(4, palette = "Dark 3")
names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
```

@@

Several authors have since pursued fully parametric approaches that made the 
fitting of smooth survival curves much more transparent and intuitive through 
generalized linear models. The key feature of these procedures is splitting the 
time axis into discrete intervals. Whitehead [-@whitehead1980fitting] showed the 
equivalence between a Cox model and a Poisson regression with a parameter for 
each event time; Carstensen [-@copenhagen2012needs] provides a nice exposition of this 
equivalence with a real data example and supporting R code for computing 
standard errors. Arjas & Haara [-@arjas1987logistic] and Efron [-@efron1988logistic] treated each 
patient-day as a Bernoulli random variable with probability equal to the 
discrete hazard rate. A potential issue with these approaches is that the 
number of time bins need to be chosen by the data analyst. On the one hand, a fine grouping 
of times may result in few (or none) events per interval, which then leads 
to instability in the Newton-Raphson procedure for estimation 
[@kalbfleisch2011statistical]. On the other hand, a coarse grouping could potentially mask 
nonlinear trends in the hazard function.

Rather than discretizing time, Hanley & Miettinen [-@hanley2009fitting] 
selected a discrete set of person-time coordinates ("person-moments") in 
continuous time from all observed follow-up experience constituting the 
study base. By doing so, they obtained a likelihood expression for the hazard 
function that is equivalent to that of logistic regression. More specifically, all person-moments when the event of interest 
occurred are selected as the case series, complemented by a randomly sampled 
base series of person-moments serving as controls. This approach allows 
flexible modeling of the hazard function by including time as a covariate 
(e.g. using splines or general additive models). Furthermore, time-dependent 
covariates can be modeled through interactions with time. In short, Hanley & 
Miettinen [-@hanley2009fitting] use the well-understood logistic regression 
for directly modeling the hazard function, without requiring a discrete-time model.

@@

```{r erspc-data-mutate, eval = eval_cs1, echo = FALSE}
data("ERSPC")
# ERSPC$ScrArm was turned to factor in 0.10.1
if (utils::packageVersion("casebase") < package_version("0.10.1")) {
  ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
                       levels = c(0,1), 
                       labels = c("Control group", "Screening group"))
}
```

```{r erspc-data, eval = eval_cs1}
pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
                               event = "DeadOfPrCa", exposure = "ScrArm")
inherits(pt_object, "popTime")
attr(pt_object, "exposure")
```

@@

For our first case study, we return to the ERSPC study and investigate the differences in risk between the control and screening arms. We fit a flexible hazard by using a smooth function of time modeled with cubic B-splines. The model is fit using \code{fitSmoothHazard} with the familiar formula interface:

```{r, eval = eval_cs1, echo = -1}
set.seed(12)
library(splines)
fit <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
                       data = ERSPC, ratio = 100)
```

The output object from \code{fitSmoothHazard} inherits from the \code{singleEventCB} and \code{glm} classes. For this reason, we can leverage the \code{summary} method for \code{glm} objects to output a familiar summary of the results:

@@

@@

```{r, eval=eval_cs1, echo=echo_plot_code}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- c(0, 1)

@@

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated risk function from \pkg{casebase} on the Cox model estimate. As noted in the Theoretical Details section, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models. Therefore, we can easily test the significance of the spline model:

```{r erscp-compare, eval = eval_cs1, echo = FALSE}
anova(fit, test = "LRT")
```

The null hypothesis is rejected in favor of the spline model. Similarly, to compare different models (e.g. time modeled linearly), we could compute Akaike's Information Criterion (AIC) for each model and compare them.

@@

```{r erspc-cox-cif, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk functions for control and screening groups in the ERSPC data. We plot the estimate from the Cox model using \\code{survival::survfit} (solid line) and the estimate from case-base sampling (dashed line). We used a cubic B-spline expansion of time."}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
myCols <- cbPalette[c(4,7)]

@@

Next, we can compute confidence intervals for the risk function using the method \code{confint.absRiskCB}.

```{r eval = eval_cs1, echo = -1}
new_time <- c(0, 1)
risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
conf_ints <- confint(risk, fit)
conf_ints
```

These point-wise confidence intervals are constructed using parametric bootstrap; the validity of this approach relies on the asymptotic normality of the MLE. In Figure \ref{fig:erspc-cif-conf}, we see the 95% confidence bands around the estimates.

```{r erspc-cif-conf, eval = eval_cs1, fig.align='h', echo = FALSE, fig.width=8, fig.height=6, fig.cap="Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\\% confidence bands."}
# confidence interval
tt <- confint(risk_mp, fit)

# plot absrisk and 95% CI
new_time <- seq(0,14,0.1)
i.backw <- order(new_time, decreasing = TRUE)
i.forw <- order(new_time)

@@

Previous re-analyses of these data suggest that the 20% reduction in prostate cancer death due to screening was an underestimate [@hanley2010mortality]. The estimated 20% (from a proportional hazards model) did not account for the delay between screening and the time the effect is expected to be observed. As a result, the null effects in years 1–7 masked the substantial reductions that began to appear from year 8 onward. This motivates the use of a time-dependent hazard ratio which can easily be fit with the \pkg{casebase} package by including an interaction term with time in the model:

```{r, echo=-1, eval = eval_cs1}
set.seed(12)
fit_inter <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) * ScrArm, 
                             data = ERSPC)
```

Note that we did not have to specify any cut points, as would be the case with the \code{survSplit} function in the \pkg{survival} package. In Figure \ref{fig:interaction-ERSPC}, we have the estimated hazard ratio and 95% confidence interval for screening vs. control group as a function of time using the \code{plot} method for objects of class \code{singleEventCB}:

```{r, echo = TRUE, eval = FALSE}
plot(fit_inter, type = "hr", newdata = new_data,
     var = "ScrArm", xvar = "Follow.Up.Time", ci = TRUE)
```

```{r interaction-ERSPC, eval = eval_cs1, fig.width=8, fig.height=6, fig.cap='Estimated hazard ratio and 95\\% confidence interval for screening vs. control group as a function of time in the ERSPC dataset. Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic B-spline basis of follow-up time and treatment arm. 95\\% confidence intervals are calculated using the delta method. The plot shows that the effect of screening only begins to become statistically apparent by year 7. The 25-60\\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\\% reported in the original article.'}
new_time <- seq(1, 12, by  = 0.1)
new_data <- data.frame(ScrArm = factor("Control group",
                                         levels = c("Control group","Screening group")),
                      Follow.Up.Time = new_time)

par(oma = c(0,1,0,8))
tt <- plot(fit_inter,
     type = "hr",
     newdata = new_data,
     var = "ScrArm",
     increment = 1,
     xvar = "Follow.Up.Time",
     ci = T,
     xlab = "Follow-up time (years)",
     ylab = "Death from prostate cancer hazard ratio",
     ylim = c(0,1.50),
     yaxt = 'n',
     xaxt = "n",
     rug = TRUE)

@@

We want to compare our hazard ratio estimates to that obtained from a Cox regression.

@@

```{r supportData, eval = eval_cs3}
scaleFUNy <- function(x) sprintf("%.2f", x)
scaleFUNx <- function(x) sprintf("%.0f", x)
data(support)
# Change time to years
support$d.time <- support$d.time/365.25

@@

We would like to thank the anonymous reviewers for their insightful comments and criticisms. We would also like to thank Yi Yang for helpful discussions on penalized regression models. \mbox{Bhatnagar} and \mbox{Turgeon} both gratefully acknowledge funding via a Discovery Grant from the Natural Sciences and Engineering Research Council of Canada (NSERC).

### 2021-10-24 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
add NSERC DG in acknowledgment

Rather than discretizing time, Hanley \& Miettinen
\citeyearpar{hanley2009fitting} selected a discrete set of person-time
coordinates (`person-moments') in continuous time from all observed
follow-up experience constituting the study base. By doing so, they
obtained a likelihood expression for the hazard function that is
equivalent to that of logistic regression. More specifically, all
person-moments when the event of interest occurred are selected as the
case series, complemented by a randomly sampled base series of
person-moments serving as controls. This approach allows flexible
modeling of the hazard function by including time as a covariate
(e.g.~using splines or general additive models). Furthermore,
time-dependent covariates can be modeled through interactions with time.
In short, Hanley \& Miettinen \citeyearpar{hanley2009fitting} use the
well-understood logistic regression for directly modeling the hazard
function, without requiring a discrete-time model.

>>>>

Rather than discretizing time, Hanley \& Miettinen
\citeyearpar{hanley2009fitting} selected a discrete set of person-time
coordinates (``person-moments'') in continuous time from all observed
follow-up experience constituting the study base. By doing so, they
obtained a likelihood expression for the hazard function that is
equivalent to that of logistic regression. More specifically, all
person-moments when the event of interest occurred are selected as the
case series, complemented by a randomly sampled base series of
person-moments serving as controls. This approach allows flexible
modeling of the hazard function by including time as a covariate
(e.g.~using splines or general additive models). Furthermore,
time-dependent covariates can be modeled through interactions with time.
In short, Hanley \& Miettinen \citeyearpar{hanley2009fitting} use the
well-understood logistic regression for directly modeling the hazard
function, without requiring a discrete-time model.
